SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2018-12-14 scipy-1.0/paper.tex
Minor revisions

Despite its success, SciPy still faces challenges:
a low ``bus factor''---from the phrase ``hit by a bus'' and
meaning that the project depends on a small number of
specific contributors---being perhaps the most serious.

@@

The availability of a standard numerical array data structure in
Python led to a sudden and rapidly growing number of scientific
packages becoming available for solving various common numeric
problems.
A number of these packages were written by graduate students and
postdoctoral researchers to solve the very practical research problems
that they faced on a daily basis.  While they had access to and were
familiar with specialized (often commercial) systems, many found
it easier to implement the domain-specific functionality they needed
in a general purpose programming language.  And given Python's innate
ability to function as a systems language, controlling specialized or
custom-built hardware was feasible.

@@

% PhD Mayo Clinic, started faculty position at BYU in 2001
By the late 1990s, discussions appeared on Matrix-SIG,
expressing a desire for a complete scientific data analysis environment\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
Travis Oliphant, a PhD student at the Mayo Clinic,
released a number of packages\footnote{https://web.archive.org/web/19990125091242/http://oliphant.netpedia.net:80/}\footnote{https://web.archive.org/web/20001206213500/http://oliphant.netpedia.net:80/} that built on top of the Numeric array
package, and provided algorithms for signal processing, special
functions, sparse matrices, quadrature, optimization, and Fast Fourier
Transforms, and more.  One of these packages, Multipack, was a set of
extension modules that wrapped Fortran and C libraries such as
ODEPACK, QUADPACK, and MINPACK, to solve and minimize nonlinear
equations, integrate differential equations, and fit splines (the
latter implemented by Pearu Peterson).  Robert Kern, then an
undergraduate student (and currently a SciPy core developer), provided
compilation instructions under Windows.
% 2001 PhD Tallinn Technical University
Around the same time, Pearu Peterson, a PhD student from Estonia,
released f2py\footnote{https://mail.python.org/pipermail/numpy-discussion/2000-September/000243.html}, a command line tool for binding Python and Fortran
codes, and wrote modules for linear algebra and interpolation.
% Graduated 1999, Duke
Eric Jones, while a graduate student at Duke, wrote a number of
packages to support his dissertation, including a parallel job
scheduler and genetic optimizer.
% Graduated 1998, PhD from Brown, 97–99 postdoc at Harvard Medical
% School; at Harvard ever since
%
% https://mail.python.org/pipermail/scipy-dev/2001-November/000088.html
%
Gary Strangman, a postdoctoral fellow at Harvard Medical School,
published several descriptive and inferential statistical routines\footnote{https://web.archive.org/web/20001022231108/http://www.nmr.mgh.harvard.edu/Neural\_Systems\_Group/gary/python.html}.

@@

As SciPy, the algorithms library of the ecosystem, began attracting the attention
of large research organizations,
the next generation of graduate students and postdocs were already exploring
other aspects of the computational environment.
In 2000, Prabhu Ramachandran\fixme{also note that he, and the others
  mentioned below, are also SciPy contributors}, a PhD student at the Indian Institute of
Technology, started work on a 3D visualization application, based on
Wingware's C++ Visualization Toolkit, called
Mayavi\footnote{https://web.archive.org/web/20030731122452/http://pythonology.org/success\&story=mayavi}.
In 2001, Fernando Pérez, a graduate student at the University of
Colorado, Boulder, created the IPython interactive shell, built into
IPP, an earlier attempt by Janko Hauser, and including code from
Nathan Gray's LazyPython.
% IPython hosted on SciPy:
% https://mail.python.org/pipermail/scipy-user/2003-March/001455.html
John Hunter, a graduate student
\fixme{check, since John around that time became postdoc, can check
  with John Milton?} at the University
of Chicago Medical School, liked the plotting functionality
available in MATLAB, but had problems accessing their laboratory's
license, which was governed by a hardware dongle.  In response, he
wrote a plotting library from scratch, and Matplotlib 0.1 was released
April 2003\footnote{https://mail.python.org/pipermail/python-list/2003-April/193167.html}.

@@

In April 2006, Paul
Dubois\footnote{https://mail.python.org/pipermail/numpy-discussion/2006-April/007487.html} wrote
to the NumPy mailing list:
\begin{quote}
IEEE's magazine, Computing in Science and Engineering (CiSE), has asked me
to put together a theme issue on the use of Python in Science and
Engineering. I will write an overview to be accompanied by 3-5 articles [. . .]
that show a diverse set of applications or
tools, to give our readers a sense of whether or not Python might be useful
in their own work.
\end{quote}
In the resulting CiSE May/June 2007 special issue, Paul's overview\cite{dubois2007guest}
stated:
\begin{quote}
LLNL now has many Python-based efforts built from scratch or wrapped around
legacy codes [. . .] hundreds of thousands
of lines of C++, Python, and Fortran 95, all working together just as we hoped,
doing compute-intensive calculations on massively parallel computers.
\end{quote}
While it is now more commonly accepted, Paul took the time to explain to
the 2007 reader that ``interpreted doesn't mean slow or only interactive.''
He also shares his own interest in Python for ``computational steering''
where ``Python serves as the input language to a scientific application, and the actual
computations are performed both in Python itself and in compiled extensions.''
%https://www.computer.org/csdl/mags/cs/2007/03/c3007.html
Using Python for computational steering was one of the earliest uses of Python
in large scientific processing pipelines.
In addition to Paul's overview there were also articles introducing
Python and SciPy \cite{oliphant2007python}, IPython \cite{perez2007ipython},
and  Matplotlib \cite{hunter2007matplotlib}.
In addition to these longer introductory articles, there was a diverse set of
research applications including
systems biology\cite{myers2007python},
astronomy\cite{greenfield2007reaching},
robotics\cite{krauss2007python},
nanophotonics\cite{bienstman2007python},
partial differential equations\cite{mardal2007using},
neuroimaging\cite{millman2007analysis},
geographic information systems\cite{shi2007python}, and
education\cite{backer2007computational, myers2007pythona}.

@@

By the middle to late 2000s, SciPy matured after a long phase of significant
growth and adoption.
The informal workshops are grown into international conferences with many
hundreds of attendees, special issues are organized and published in a
leading scientific magazine, and the scope of the SciPy library
narrows, while the breadth of the ecosystem is grown and encouraged
through a new type of auxiliary package, the scikit.  The tooling,
development and release processes become more professional, grown with
a patience only affordable in open source development, leading with
best practices nowadays increasingly common in both industry and other
open source projects \cite{millman2014developing}.

@@

In the early workshops, recurrent topics indicate the state of development, with
emphasis being placed on the underlying array package, plotting,
parallel processing, acceleration / wrapping, and user interfaces.  By
2004, a significant shift occurred towards application of SciPy on
scientific problems.  The event also started to draw in more keynote
speakers from outside the community, such as Guido van Rossum (creator
of Python, 2006), Ivan Krstić (One Laptop per Child, 2007), Alex
Martelli (Google, 2008), and Peter Norvig (Google Research, 2009).
The SciPy conference went from being a small gathering of
core developers, to a multi-location event, with increased funding, a
published proceedings, and scholarships for attending students (some
of whom are authors on this paper).  In 2010, the conference grew even
further, with an additional event added in India, and the US
conference—now with parallel tracks—moving to Austin, Texas.
While the US SciPy conference kept growing (by ... it had multiple
tracks, student sponsorship, proceedings, and funding), satellite
events were being organized by volunteers elsewhere, such as EuroSciPy
(2008–) and SciPy India (2009–).
Special sessions and minisymposia dedicated to scientific Python began
appearing at many events.
For example, a three-part minisymposium organized for CSE 2009 was
featured in SIAM News\cite{siamcse09}.

@@

Python has become so pervasive, there are now domain-specific special issues.
For example, Frontiers in Neuroinformatics has a collection of 25 articles---covering
topics including modeling and simulation, data collection, electrophysiology, visualization,
as well as stimulus generation and presentation---called Python in
neuroscience\footnote{\url{https://www.frontiersin.org/research-topics/8/python-in-neuroscience}}

In 2012, Perry Greenfield, John Hunter, Jarrod Millman, Travis Oliphant,
and Fernando Pérez founded NumFOCUS\footnote{https://numfocus.org}.
NumFOCUS is a 501(c)3 public charity whose mission is to
``promote sustainable high-level programming languages, open code development,
and reproducible scientific research.''
While NumFOCUS is language agnostic, many of the early sponsored projects
came from the scientific Python stack.
Today it has over 25 sponsored projects including NumPy, SciPy, IPython, and
Matplotlib.
Among its other projects, NumFOCUS organizes a global network of 
community driven educational program called PyData.

>>>>

Despite its success, SciPy still faces challenges:
a low ``bus factor''---meaning that the entire project depends on a small number of
active contributors---being perhaps the most serious.

@@

The availability of a standard numerical array data structure in
Python led to a sudden and rapidly growing number of scientific
packages becoming available for solving various common numeric
problems.
A number of these packages were written by graduate students and
postdoctoral researchers to solve the very practical research problems
that they faced on a daily basis.  While they had access to and were
familiar with specialized (often commercial) systems, many found
it easier to implement the domain-specific functionality they needed
in a general purpose programming language.  And given Python's innate
ability to function as a systems language, controlling specialized or
custom-built hardware was also possible.

@@

% PhD Mayo Clinic, started faculty position at BYU in 2001
By the late 1990s, discussions appeared on Matrix-SIG,
expressing a desire for a complete scientific data analysis environment\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
Travis Oliphant, a PhD student at the Mayo Clinic,
released a number of packages\footnote{https://web.archive.org/web/19990125091242/http://oliphant.netpedia.net:80/}\footnote{https://web.archive.org/web/20001206213500/http://oliphant.netpedia.net:80/} that built on top of the Numeric array
package, and provided algorithms for signal processing, special
functions, sparse matrices, quadrature, optimization, Fast Fourier
Transforms, and more.  One of these packages, Multipack, was a set of
extension modules that wrapped Fortran and C libraries such as
ODEPACK, QUADPACK, and MINPACK, to solve and minimize nonlinear
equations, integrate differential equations, and fit splines (the
latter implemented by Pearu Peterson).  Robert Kern, then an
undergraduate student (and currently a SciPy core developer), provided
compilation instructions under Windows.
% 2001 PhD Tallinn Technical University
Around the same time, Pearu Peterson, a PhD student from Estonia,
released f2py\footnote{https://mail.python.org/pipermail/numpy-discussion/2000-September/000243.html}, a command line tool for binding Python and Fortran
codes, and wrote modules for linear algebra and interpolation.
% Graduated 1999, Duke
Eric Jones, while a graduate student at Duke, wrote a number of
packages to support his dissertation, including a parallel job
scheduler and genetic optimizer.
% Graduated 1998, PhD from Brown, 97–99 postdoc at Harvard Medical
% School; at Harvard ever since
%
% https://mail.python.org/pipermail/scipy-dev/2001-November/000088.html
%
Gary Strangman, a postdoctoral fellow at Harvard Medical School,
published several descriptive and inferential statistical routines\footnote{https://web.archive.org/web/20001022231108/http://www.nmr.mgh.harvard.edu/Neural\_Systems\_Group/gary/python.html}.

@@

As SciPy, the algorithms library of the ecosystem, began attracting the attention
of large research organizations,
the next generation of graduate students and postdocs were already exploring
other aspects of the computational environment.
In 2000, Prabhu Ramachandran, a PhD student at the Indian Institute of
Technology, started work on a 3D visualization application, based on
Wingware's C++ Visualization Toolkit, called
Mayavi\footnote{https://web.archive.org/web/20030731122452/http://pythonology.org/success\&story=mayavi}.
In 2001, Fernando Pérez, a graduate student at the University of
Colorado, Boulder, created the IPython interactive shell (that would
eventually blossom into Project Jupyter), built into
IPP, an earlier attempt by Janko Hauser, and including code from
Nathan Gray's LazyPython.
% IPython hosted on SciPy:
% https://mail.python.org/pipermail/scipy-user/2003-March/001455.html
John Hunter, a postdoc
%John got his PhD in 2001, then a post-doc with John Milton (dept of
%Neurology, UChicago?) a couple of years; then research professor
%under Kurt Hecox, whereafter at Tradelink from 2006
at the University
of Chicago, liked the plotting functionality
available in MATLAB, but had problems accessing their laboratory's
license, which was governed by a hardware dongle.  In response, he
wrote a plotting library from scratch, and Matplotlib 0.1 was released
April 2003\footnote{https://mail.python.org/pipermail/python-list/2003-April/193167.html}.

@@

%In April 2006, Paul
%Dubois\footnote{https://mail.python.org/pipermail/numpy-discussion/2006-April/007487.html} wrote
%to the NumPy mailing list:
%\begin{quote}
%IEEE's magazine, Computing in Science and Engineering (CiSE), has asked me
%to put together a theme issue on the use of Python in Science and
%Engineering. I will write an overview to be accompanied by 3-5 articles [. . .]
%that show a diverse set of applications or
%tools, to give our readers a sense of whether or not Pyth%on might be useful
%in their own work.
%\end{quote}
In a May/June 2007 special issue of IEEE Computing in Science and
Engineering, Paul Dubois wrote\cite{dubois2007guest}:
\begin{quote}
LLNL now has many Python-based efforts built from scratch or wrapped around
legacy codes [. . .] hundreds of thousands
of lines of C++, Python, and Fortran 95, all working together just as we hoped,
doing compute-intensive calculations on massively parallel computers.
\end{quote}
While it is now more commonly accepted, Paul took the time to explain to
the 2007 reader that ``interpreted doesn't mean slow or only interactive.''
He also shares his own interest in Python for ``computational steering''
where ``Python serves as the input language to a scientific application, and the actual
computations are performed both in Python itself and in compiled extensions.''
%https://www.computer.org/csdl/mags/cs/2007/03/c3007.html
Using Python for computational steering was one of the earliest uses of Python
in large scientific processing pipelines.
In addition to Paul's overview there were also articles introducing
Python and SciPy \cite{oliphant2007python}, IPython \cite{perez2007ipython},
and  Matplotlib \cite{hunter2007matplotlib}.
In addition to these longer introductory articles, there was a diverse set of
research applications including
systems biology\cite{myers2007python},
astronomy\cite{greenfield2007reaching},
robotics\cite{krauss2007python},
nanophotonics\cite{bienstman2007python},
partial differential equations\cite{mardal2007using},
neuroimaging\cite{millman2007analysis},
geographic information systems\cite{shi2007python}, and
education\cite{backer2007computational, myers2007pythona}.

@@

By the middle to late 2000s, SciPy was starting to mature after a long phase of significant
growth and adoption.
The informal workshops are grown into international conferences with many
hundreds of attendees, special issues are organized and published in a
leading scientific magazine, and the scope of the SciPy library
narrows, while the breadth of the ecosystem is grown and encouraged
through a new type of auxiliary package, the scikit.  The tooling,
development, and release processes become more professional, grown with
the patience affordable in open source development, leading with
best practices nowadays increasingly common in both industry and other
open source projects \cite{millman2014developing}.

@@

In the early workshops, recurrent topics indicate the state of development, with
emphasis being placed on the underlying array package, plotting,
parallel processing, acceleration / wrapping, and user interfaces.  By
2004, a significant shift occurred towards application of SciPy to
scientific problems.  The event also started to draw in more keynote
speakers from outside the community, such as Guido van Rossum (creator
of Python, 2006), Ivan Krstić (One Laptop per Child, 2007), Alex
Martelli (Google, 2008), and Peter Norvig (Google Research, 2009).
The SciPy conference went from being a small gathering of
core developers, to a multi-location event, with increased funding, a
published proceedings, and scholarships for attending students (some
of whom are authors on this paper).  In 2010, the conference grew even
further, with an additional event added in India, and the US
conference—now with parallel tracks—moving to Austin, Texas.
While the US SciPy conference kept growing (by ... it had multiple
tracks, student sponsorship, proceedings, and funding), satellite
events were being organized by volunteers elsewhere, such as EuroSciPy
(2008–) and SciPy India (2009–).
Special sessions and minisymposia dedicated to scientific Python began
appearing at many events.
For example, a three-part minisymposium organized for CSE 2009 was
featured in SIAM News\cite{siamcse09}.

@@

Python has become so pervasive, there are now domain-specific special issues.
For example, Frontiers in Neuroinformatics has a collection of 25 articles---covering
topics including modeling and simulation, data collection, electrophysiology, visualization,
as well as stimulus generation and presentation---called Python in
neuroscience\footnote{\url{https://www.frontiersin.org/research-topics/8/python-in-neuroscience}}.

In 2012, Perry Greenfield, John Hunter, Jarrod Millman, Travis Oliphant,
and Fernando Pérez founded NumFOCUS\footnote{https://numfocus.org},
 a 501(c)3 public charity whose mission is to
``promote sustainable high-level programming languages, open code development,
and reproducible scientific research.''
While NumFOCUS is language agnostic, many of the early sponsored projects
came from the scientific Python stack.
Today it has over 25 sponsored projects including NumPy, SciPy, IPython, and
Matplotlib.
Among its other projects, NumFOCUS organizes a global network of 
community driven educational program called PyData.
